---
title: "2016 Presidential Election Finances (DAN DAR/P4 FINAL PROJECT)"
author: "Andri Rizhakov"
date: "February 5, 2016"
---
***
***

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, 
                      fig.height=6, 
                      fig.path='Figs/', 
                      echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE
                      )
```

```{r packages}
## load the libraries
library(ggplot2)
library(GGally)
library(scales) 
library(memisc)
library(gridExtra)
library(dplyr)
library(reshape2)
```

```{r Load_the_Data}
### Load the Data
## set correct directory
setwd("C:/Users/arizhako/Box Sync/Reference/programming/python/DAN/P4_DAR")
#setwd("C:/Users/Andri/Google Drive/Dev/mind/courses/udacity_DAN/P4_DAR/FINAL")
```

# 1. Exploratory Data Analysis: 2016 Presidential Campaign Finance data set

The 2016 Presidential Campaign Finance data set was chosen because of personal 
interest and special relevance when doing this project (i.e., it is the 
election year!). 

No particular questions for analysis come to mind, but I hope they will arise 
as I play around and explore and understand the data set. Generally, I am 
curious to learn about the demographics of who (i.e., which states, 
characterstics of donators, etc.) is donating to the various campaigns, 
how much, and to whom. Additionally, any insight inferred from who is winning 
the 2016 presidential race, given the financing information, would be 
interesting, and would make watching the elections even more exciting (to see 
how close I was!). 

Raw data sets were downloaded from the FEC [Ref. 1].

_**Part 1.1. Basic import and explore**_
```{r 1_1}
## import main dataset 
prez2016 <- read.csv('P00000001-ALL_2016.csv',
                     row.names = NULL)
colnames(prez2016) <- c(colnames(prez2016)[-1],"x")
prez2016$x <- NULL

# dataset structure summary
str(prez2016)
```

Results give an indication of the various variable types and the number of 
unique field values. Most variables are Factor type. For the needs of this 
analysis, some fields are not necessary and will be carried through in the 
dataframes but mostly ignored (e.g., "cmte\_id", "tran\_id").

_**Part 1.2. Summary**_
```{r 1_2}
summary(prez2016)
```

Some more insights are obtained about the 2016 financing: Clinton has the most 
contribution counts (84,282 counts from "cand\_nm"); Matt Lenell has donated 
the most number of times (281 counts from "contbr\_nm"); The most donators 
came from New York City (10,478 counts from "contbr\_city"), though the most 
donators came from California (56,764 counts from "contbr\_st"). The most 
common donator is Retired (96,007 counts in "contbr\_occupation"). The mean 
contribution amount is $436, though median is $100 (from "contb\_receipt\_amt").  

_**Part 1.3. Scatterplot Matrix**_
```{r 1_3, fig.height=12}
### create scatterplot matrix
## sample 10,000 entries from the data set
set.seed(2016)

## use subset of full prez set
prez2016_trunc1 <- subset(prez2016, select = c(cand_id,
                                     cand_nm,
                                     contbr_st,
                                     contb_receipt_amt))
prez2016_trunc1.samp <- prez2016_trunc1[sample(
                                          1:length(prez2016_trunc1$cand_id), 
                                          10000), 
                                        ]

#generic 'ggpairs' implementation, can finesse if needed later
ggpairs(prez2016_trunc1.samp)

```

The scatter plot matrix gives us some preliminary plotting of the sampled data 
set and helps us understand the full data set. The structure, values, shapes of
plots, distributions, and preliminary relationships are more understood by 
plotting each variable against all others. The counts of each variable are 
plotted on the diagonal. For the off-diagonal plots, the only potentially 
useful plots are the box plots for the "contb\_receipt\_amt", because it is 
the only quantitative variable. 

_**Part 1.4. Basic plots: counts of contb receipt amt and count of each.**_
```{r 1_4}
pBar_1_2016 <- ggplot(data = prez2016, 
                      aes(x = cand_nm)) +
  geom_bar() +
  ggtitle("Contribution Count, by Candidate (2016)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

grid.arrange(pBar_1_2016)

pLine_1_2016 <- ggplot(data = subset(
                                subset(prez2016, contb_receipt_amt>0),
                                !is.na(contb_receipt_amt)
                                    ),
                       aes(x=contb_receipt_amt, 
                           y=..count..)
                       ) +
  geom_freqpoly(binwidth = 0.1) +
  ggtitle("Contribution Count, by Amount (2016)") +
  scale_x_log10(limits = c(1,10000))

grid.arrange(pLine_1_2016)

```

The basic univariate plots show the count of contributions made to each 
candidate, and the count of all contribution amounts. From the bar graph, 
Clinton, Sanders, and Carson are the three top contribution count receivers, 
with upwards of 60,000 contributions. From the line plot, it appears that the 
distribution of contribution amounts is approximately normal, after 
transforming the x-axis to logarithmic. Also, the highest contribution counts
are for approximately $100 in size.

_**Part 1.5. dplyr ops: get total contb receipt amt**_  
Create a new data frame to get total "contb\_receipt\_amt" for each 2016 
presidential candidate. The new data frame has the following structure:
```{r 1_5}
prez2016.cont_by_name <- prez2016 %>%
  group_by(cand_nm) %>%
  summarise(sum_amnts = sum(contb_receipt_amt),
            n = n()) %>%
  arrange(cand_nm)

str(prez2016.cont_by_name)
```

The total itemized contributions for each candidate are:
```{r }
###manual check
by(prez2016$contb_receipt_amt, prez2016$cand_nm, sum) 
  #match with Excel for Bernie, Rubio
```

With the sum for all candidates:
```{r }
sum(prez2016$contb_receipt_amt) #match with Excel
```

_**Part 1.6. More plots: distributions and sums of It. contribution amounts**_
```{r 1_6, fig.height=14 }
pBox_1_2016 <- ggplot(data = prez2016, 
                      aes(x = cand_nm, 
                          y = contb_receipt_amt)
                      ) +
  geom_boxplot() + 
  ylim(0, quantile(prez2016$contb_receipt_amt,0.9999)) + 
  ggtitle("Itemized Contribution Amount, by Candidate (2016)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

pBar_2_2016 <- ggplot(data = prez2016.cont_by_name, 
                      aes(x = cand_nm,
                          y = sum_amnts)) +
  geom_bar(stat="identity") + 
  ggtitle("Sum of Itemized Contributions, by Candidate (2016)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))


#plot both 
grid.arrange(pBar_1_2016, pBox_1_2016, pBar_2_2016)
```
The first bar graph of each candidates total number of contributions is 
presented along with the box plot of the same candidate and the distribution of
the contribution amounts. Additionally, from the data in the recently created
data frame shows a bar graph of the total contribution amount each candidate
has received.

Focusing on the larger amounts and larger contribution totals, Carson, Clinton, 
Cruz, and Sanders all have high contribution counts. However, Carson, Cruz, and 
Sanders have a very similar box plot distribution - they are all near $0 
contribution amount (probably more like $100, from the previous line plot). 
Clinton differs by having higher contribution amounts. Furthermore, Clinton 
has by far the highest total contribution amount. What is the cause of this 
discrepancy? 

Discrepancy found in the itemized contributions: why so much total dollar 
contribution for Clinton? Is this reality, or a bug? What is the explanation 
behind this - especially given the news that Sanders has been setting 
campaign funding records?

After emailing the FEC [Campaign finance reports and data: pubrec@fec.gov], 
then calling [phone number 8004249530, option #6], I finally spoke to a FEC 
data representative who explained this finding. I was able to go to [Ref. 2] 
for full information on the candidate. Discrepancy was due to the breakdown in 
itemized/unitemized contributions, and the *.csv datasets that are downloaded 
from the website are only for itemized contributions. For example, $9MM for 
Sanders was itemized, but he received more from unitemized - these are the 
fundraising records that were being broken. 

Next, I checked for internal consistency and that was passed. I briefly took 
the amounts that were reported in the graphic on [Ref. 1] and compared to 
manually adding the itemized and unitemized amounts from [Ref. 2]. Although 
I was not able to align perfectly to the single dollar digit, the comparison 
was much improved with an error of perhaps a few percent. For example, Clinton 
had $63,366,253.00 from Itemized Individual Contributions from .csv file total 
[Ref.3], $61,726,374 from ALL.zip .csv via R and used in this analysis 
[Ref. 1], $62,785,473 from [Ref. 2] summary page - all more or less the same, 
at least for the purposes of this project's investigation. Thusly, the 
understanding is gained for the dataset in the importance of itemized vs. 
unitemized total amounts for contribution amounts.

_**Part 1.7. Add non-itemized contributions as bulk amount**_
```{r 1_7 }
### add non-itemized contributions
## source: http://fec.gov/fecviewer/CandidateCommitteeDetail.do

prez2016.contTotal_by_name <- data.frame(prez2016.cont_by_name)
names(prez2016.contTotal_by_name)[
  names(prez2016.contTotal_by_name)=="sum_amnts"] <- "Itemized"
prez2016.contTotal_by_name["Unitemized"] <- 0
#str(prez.contTotal_by_name) ##review
#table(prez2016.contTotal_by_name$cand_nm) ##review
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Bush, Jeb", 
  "Unitemized"] <- 1244623
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Carson, Benjamin S.", 
  "Unitemized"] <- 19641649
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Christie, Christopher J.", 
  "Unitemized"] <- 141428
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Clinton, Hillary Rodham", 
  "Unitemized"] <- 13292382
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Cruz, Rafael Edward 'Ted'", 
  "Unitemized"] <- 11064064
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Fiorina, Carly", 
  "Unitemized"] <- 3980707
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Graham, Lindsey O.", 
  "Unitemized"] <- 200049
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Huckabee, Mike", 
  "Unitemized"] <- 1314344
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Jindal, Bobby", 
  "Unitemized"] <- 134279
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Kasich, John R.", 
  "Unitemized"] <- 518444
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Lessig, Lawrence", 
  "Unitemized"] <- 0
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "O'Malley, Martin Joseph", 
  "Unitemized"] <- 226725
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Pataki, George E.", 
  "Unitemized"] <- 24218
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Paul, Rand", 
  "Unitemized"] <- 4491348
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Perry, James R. (Rick)", 
  "Unitemized"] <- 139408
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Rubio, Marco", 
  "Unitemized"] <- 3257245
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Sanders, Bernard", 
  "Unitemized"] <- 30652976
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Santorum, Richard J.", 
  "Unitemized"] <- 216875
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Trump, Donald J.", 
  "Unitemized"] <- 2816148
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Walker, Scott", 
  "Unitemized"] <- 2652470
prez2016.contTotal_by_name[
  prez2016.contTotal_by_name$cand_nm == "Webb, James Henry Jr.", 
  "Unitemized"] <- 290305

prez2016.contTotal_by_name["Ratio_U_I"] <- 
  prez2016.contTotal_by_name$Unitemized / prez2016.contTotal_by_name$Itemized
prez2016.contTotal_by_name["Total"] <- 
  prez2016.contTotal_by_name$Unitemized + prez2016.contTotal_by_name$Itemized

```
Manually adding the amounts of unitemized contributions manually to a new data 
frame, we can then get a better picture for the cash flow coming into each 
candidate from most campaign sources. New data frame now has unitemized 
contribution totals from [Ref. 2] included.

_**Part 1.8. Sum of Itemized and Unitemized Contributions, by Candidate**_
```{r 1_8, fig.height=18 }
pBar_3_2016 <- ggplot(data = prez2016.contTotal_by_name, 
                      aes(x = cand_nm, y = Total)) +
  geom_bar(stat="identity") + 
  ggtitle("Sum of Itemized and Unitemized Contributions, by Candidate (2016)")+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

pBar_4_2016 <- ggplot(data = prez2016.contTotal_by_name, 
                      aes(x = cand_nm, y = Ratio_U_I)) +
  geom_bar(stat="identity") + 
  ggtitle("Ratio of Unitemized to Itemized Contributions, by Candidate (2016)")+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

#plot both 
grid.arrange(pBar_1_2016, pBox_1_2016, pBar_2_2016, pBar_3_2016, pBar_4_2016, 
             ncol=1)
```
The addition of the unitemized contributions helps paint the full picture of 
the campaign funds raised by the candidates. The addition of unitemized 
contributions data makes Sanders a true second place candidate, behind Clinton. 
Also, taking the ratio of unitemized to itemized contributions, we see that 
Sanders is the only candidate who has 3 times the unitemized contributions to 
itemized. 

_**Part 1.9. Heat map of average itemized contribution amount**_  
A new data frame is made in order to process data in creating a heat map.  
```{r 1_9, fig.height=10 }
### create a heat map: State vs. Candidate.
# organize data to long format: candidate, state, average amount, count
prez2016.cont_by_st <- prez2016 %>%
  group_by(cand_nm, contbr_st) %>%
  summarise(avg_amnts = mean(contb_receipt_amt),
            n = n()) %>%
  ungroup() %>%
  arrange(cand_nm)

#head(prez2016.cont_by_st) ##review

pHeat_1_2016 <- ggplot(aes(y = prez2016.cont_by_st$contbr_st, 
           x = prez2016.cont_by_st$cand_nm, 
           fill = prez2016.cont_by_st$avg_amnts),
       data = prez2016.cont_by_st) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(c("green", "red"))(100)) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

#plot both 
grid.arrange(pHeat_1_2016, ncol=1)
```
Heat map shows the average contribution amounts by state for each candidate. 
This can be very useful if we wanted to compare a candidate's state-averaged 
contributions to another.

To continue the exploratory data analysis, an investigation on parameter or 
parameters that can point to who is favored by America, using dollars 
contributed as their proxy vote, and if it can point to who may win the 
election.

To approach this we need to replicate above exploratory data analysis on the 
2012 and 2008 presidential data [Ref. 1] and see if there are any patterns 
that emerge. 

>> **NOTE** The use of repetitive code below may seem like a function 
contruct would be more appropriate and efficient, however, I believe that due 
to the explaratory nature of the analysis and only 2 repeats after the initial, 
a function is not warranted. Furthermore, the blocks for each year are 
slightly different and require some modification that a cooking-cutter 
approach of a function would not permit.

***

# 2. Exploratory Data Analysis: 2012 Presidential Campaign Finance data set
Purpose of this section is to see patterns with latest available data on a 
presidential electee, to eliminate inflation and other transient effects from 
analysis. Note that this is with incumbent president (Obama, assumed office 
first in 2008).

Basic import of data set, preliminary graphical exploration and investigation. 

_**Part 2.1. Basic import and explore**_
```{r 2_1}
##import 2012 dataset. 
prez2012 <- read.csv('P00000001-ALL_2012.csv', 
                     row.names = NULL)

colnames(prez2012) <- c(colnames(prez2012)[-1],"x")
prez2012$x <- NULL

#check
#head(prez2012)

# dataset structure summary
str(prez2012)

#summary(prez2012$contb_receipt_dt) ##review
```

Again, results give an indication of the various variable types and the 
number of unique field values. Most variables are Factor type. For the needs 
of this analysis, some fields are not necessary and will be carried through 
in the dataframes but mostly ignored (e.g., "cmte\_id", "tran\_id"). 

_**Part 2.2. Summary**_
```{r 2_2}
summary(prez2012)
```

Some more insights are obtained about the 2012 financing: Obama has the most 
contribution counts (4,117,404 counts from "cand\_nm"); Michael Smith has 
donated the most number of times (533 counts from "contbr\_nm"); The most 
donators came from New York City (138,289 counts from "contbr\_city"), though 
the most donators came from California (954,661 counts from "contbr\_st"). 
The most common donator is Retired (1,388,962 counts in "contbr\_occupation"). 
The mean contribution amount is $213, though median is $50 
(from "contb\_receipt\_amt"). 

_**Part 2.3. Scatterplot Matrix**_
```{r 2_3, fig.height=12}
### create scatterplot matrix
## sample 10,000 entries from the data set
set.seed(2012)

## use subset of full prez set
prez2012_trunc1 <- subset(prez2012, select = c(cand_id,
                                     cand_nm,
                                     contbr_st,
                                     contb_receipt_amt))
prez2012_trunc1.samp <- prez2012_trunc1[sample(
                                          1:length(prez2012_trunc1$cand_id),
                                          10000), 
                                        ]

#generic 'ggpairs' implementation, can finesse if needed later
ggpairs(prez2012_trunc1.samp)

```

Again, the scatter plot matrix gives us some preliminary plotting of the 
sampled data set and helps us understand the full data set. The structure, 
values, shapes of plots, distributions, and preliminary relationships are 
more understood by plotting each variable against all others. The counts of 
each variable are plotted on the diagonal. For the off-diagonal plots, the 
only potentially useful plots are the box plots for the "contb\_receipt\_amt", 
because it is the only quantitative variable. 

A detour is needed before proceeding with a comparative analysis to that of 
the 2016 Presidential data set. The main correction that needs to be applied 
is the alignment of the same period of time that was used in the 2016 
Presidential data set. The solution is to use the same 2012 Presidential 
data set, but shift in time. Specifically, because the 2016 Presidential 
data set is from 04/01/2015 to 09/30/2015 [Ref. 2], the same time frame needs 
to be applied to 2012 Presidential data set - 04/01/2011 to 09/30/2011. Thus, 
we need to use the receipt date field to convert to a date that R can sort on 
and then do the analysis on the normalized subset.

```{r }
prez2012$contb_receipt_dt <- as.Date(prez2012$contb_receipt_dt, "%d-%b-%y")
#head(prez2012$contb_receipt_dt) ##review


## the eq data is from  04/01/2011 - 09/30/2011
#summary(prez2012$contb_receipt_amt>10000) ##review
#summary(prez2012$contb_receipt_dt>"2011-04-01" & 
  #prez2012$contb_receipt_dt<"2011-09-30") ##review
prez2012.normDate <- subset(prez2012,
                            (prez2012$contb_receipt_dt>="2011-04-01" &    
                               prez2012$contb_receipt_dt<="2011-09-30")
                            )
```

_**Part 2.4. Basic plots: counts of itemized contributions and count of each.**_
```{r 2_4 }
pBar_1_2012 <- ggplot(data = prez2012.normDate, 
                 aes(x = cand_nm)) +
  geom_bar() +
  ggtitle("Contribution Count, by Candidate (2012)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

grid.arrange(pBar_1_2012)


pLine_1_2012 <- ggplot(data = subset(
                                subset(prez2012.normDate, contb_receipt_amt>0),
                                !is.na(contb_receipt_amt)
                                    ),
                       aes(x=contb_receipt_amt, 
                           y=..count..)
                       ) +
  geom_freqpoly(binwidth = 0.1) +
  ggtitle("Contribution Count, by Amount (2012)") +
  scale_x_log10(limits = c(1,10000))

grid.arrange(pLine_1_2012)
```
The basic univariate plots show the count of contributions made to each 
candidate, and the count of all contribution amounts. From the bar graph, 
Obama, Romney, and Paul are the three top contribution count receivers, with 
upwards of 25,000 contributions. Obama has the most counts by far, but this 
is probably because he is absorbing all Democrat donations due to being the 
only Democrat candidate because he is the incumbent. From the line plot, it 
appears that the distribution of contribution amounts is approximately normal, 
after transforming the x-axis to logarithmic. Also, the highest contribution 
counts are for approximately $100 in size.

_**Part 2.5. dplyr ops: get total contb receipt amt**_  
Create a new data frame to get total "contb\_receipt\_amt" for each 2012 
presidential candidate. The new data frame has the following structure:
```{r 2_5 }
prez2012.normDate.cont_by_name <- prez2012.normDate %>%
  group_by(cand_nm) %>%
  summarise(sum_amnts = sum(contb_receipt_amt),
            n = n()) %>%
  arrange(cand_nm)

str(prez2012.normDate.cont_by_name)
```

The total itemized contributions for each candidate are:
```{r  }
###manual check
by(prez2012.normDate$contb_receipt_amt, prez2012.normDate$cand_nm, sum)
```

With the sum for all candidates:
```{r  }
sum(prez2012.normDate$contb_receipt_amt)
```

_**Part 2.6. More plots: distributions and sums of It. contribution amounts**_
```{r 2_6, fig.height=14 }
pBox_1_2012 <- ggplot(data = prez2012.normDate, 
                 aes(x = cand_nm, 
                     y = contb_receipt_amt)
                 ) +
  geom_boxplot() + 
  ylim(0, quantile(prez2012.normDate$contb_receipt_amt,0.9999)) + 
  ggtitle("Itemized Contribution Amount, by Candidate (2012)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

pBar_2_2012 <- ggplot(data = prez2012.normDate.cont_by_name, 
                 aes(x = cand_nm, 
                     y = sum_amnts)
                 ) +
  geom_bar(stat="identity") + 
  ggtitle("Sum of Itemized Contributions, by Candidate (2012)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))


#plot both 
grid.arrange(pBar_1_2012, pBox_1_2012, pBar_2_2012)
#
```
The first bar graph of each candidates total number of contributions is 
presented along with the box plot of the same candidate and the distribution 
of the contribution amounts. Additionally, from the data in the recently 
created data frame shows a bar graph of the total contribution amount each 
candidate has received.

Focusing on the larger amounts and larger contribution totals, Obama, Romney, 
and Paul all have high contribution counts. However, Obama and Paul have a 
very similar box plot distribution - they are all near $0 contribution amount 
(probably more like $100, from the previous line plot). Romney differs by 
having higher contribution amounts. Furthermore, Romney has the highest total 
contribution amount. The discrepancy is again probably due to itemized vs. 
unitemized contributions, but cannot add non-itemized contributions as bulk 
amount because unitemized is lumped for the whole 2 year period. Addition of 
the full 2 year amount could potentially skew results. Therefore, Steps 7 and 
8 are skipped.

_**Part 2.9. Heat map of average itemized contribution amount **_

A new data frame is made in order to process data in creating a heat map.   
```{r 2_9,fig.height=10 }
### create a hap map: State vs. Candidate.
# organize data to long format: candidate, state, average amount, count
prez2012.normDate.cont_by_st <- prez2012.normDate %>%
  group_by(cand_nm, contbr_st) %>%
  summarise(avg_amnts = mean(contb_receipt_amt),
            n = n()) %>%
  ungroup() %>%
  arrange(cand_nm)

#head(prez2012.normDate.cont_by_st) ##review

pHeat_1_2012 <- ggplot(aes(y = prez2012.normDate.cont_by_st$contbr_st, 
           x = prez2012.normDate.cont_by_st$cand_nm, 
           fill = prez2012.normDate.cont_by_st$avg_amnts),
  data = prez2012.normDate.cont_by_st) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(c("green", "red"))(100)) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

#plot both 
grid.arrange(pHeat_1_2012)
```
Again, heat map shows the average contribution amounts by state for each 
candidate. This can be very useful if we wanted to compare a candidate's 
state-averaged contributions to another.

Proceeding further in the exploratory data analysis, an investigation on the 
2008 presidential campaign finance set is performed. This should eliminate 
Obama's incumbency status as a skewing variable in the 2012 data set.

***

# 3. Exploratory Data Analysis: 2008 Presidential Campaign Finance data set
Purpose of this section is to see patterns with a non-incumbent candidate 
(2012 was Obama's incumbency), and compare with the non-incumbent election 
of 2016. 

Basic import of data set, preliminary graphical exploration and investigation. 

_**Part 3.1. Basic import and explore**_
```{r 3_1 }
##import 2012 dataset. 
prez2008 <- read.csv('P00000001-ALL_2008.csv', 
                     row.names = NULL)

colnames(prez2008) <- c(colnames(prez2008)[-1],"x")
prez2008$x <- NULL

#check
#head(prez2008)

# dataset structure summary
str(prez2008)

#summary(prez2008$contb_receipt_dt)
```

Again, results give an indication of the various variable types and the number 
of unique field values. Most variables are Factor type. For the needs of this 
analysis, some fields are not necessary and will be carried through in the 
dataframes but mostly ignored (e.g., "cmte\_id", "tran\_id"). 

_**Part 3.2. Summary**_
```{r 3_2 }
summary(prez2008)
```

Some more insights are obtained about the 2008 financing: Obama has the most 
contribution counts (2,600,437 counts from "cand\_nm"); Doodad Pro has donated 
the most number of times (1627 counts from "contbr\_nm"); The most donators 
came from New York City (163,340 counts from "contbr\_city"), though the most 
donators came from California (712,532 counts from "contbr\_st"). The most 
common donator is Retired (784,290 counts in "contbr\_occupation"). The mean 
contribution amount is $257, though median is $100 (from "contb\_receipt\_amt").  

_**Part 3.3. Scatterplot Matrix**_
```{r 3_3, fig.height=12}
### create scatterplot matrix
## sample 10,000 entries from the data set
set.seed(2008)

## use subset of full prez set
prez2008_trunc1 <- subset(prez2008, select = c(cand_id,
                                     cand_nm,
                                     contbr_st,
                                     contb_receipt_amt))
prez2008_trunc1.samp <- prez2008_trunc1[sample(
                                              1:length(prez2008_trunc1$cand_id), 
                                              10000), 
                                              ]

#generic 'ggpairs' implementation, can finesse if needed later
ggpairs(prez2008_trunc1.samp)

```

Again, the scatter plot matrix gives us some preliminary plotting of the 
sampled data set and helps us understand the full data set. The structure, 
values, shapes of plots, distributions, and preliminary relationships are 
more understood by plotting each variable against all others. The counts of 
each variable are plotted on the diagonal. For the off-diagonal plots, the 
only potentially useful plots are the box plots for the "contb\_receipt\_amt", 
because it is the only quantitative variable. 

Again, a detour is needed before proceeding with a comparative analysis to 
that of the 2016 Presidential data set. The main correction that needs to be 
applied is the alignment of the same period of time that was used in the 2016 
Presidential data set. The solution is to use the same 2008 Presidential data 
set, but shift in time. Specifically, because the 2016 Presidential data set 
is from 04/01/2015 to 09/30/2015, the same time frame needs to be applied to 
2008 Presidential data set - 04/01/2007 to 09/30/2007. Thus, we need to use 
the receipt date field to convert to a date that R can sort on and do the 
analysis on the normalized subset.

```{r  }
#as.Date(mydate2, "%d-%b-%y")
prez2008$contb_receipt_dt <- as.Date(prez2008$contb_receipt_dt, "%d-%b-%y")
#head(prez2008$contb_receipt_dt)


## the eq data is from  04/01/2011 - 09/30/2011
#summary(prez2008$contb_receipt_amt>10000)
#summary(prez2008$contb_receipt_dt>"2007-04-01" & 
  #prez2008$contb_receipt_dt<"2007-09-30")
prez2008.normDate <- subset(prez2008,
                            (prez2008$contb_receipt_dt>="2007-04-01" &    
                               prez2008$contb_receipt_dt<="2007-09-30")
                            )

```

_**Part 3.4. Basic plots: counts of itemized contributions and count of each.**_
```{r 3_4 }
pBar_1_2008 <- ggplot(data = prez2008.normDate, 
                 aes(x = cand_nm)) +
  geom_bar() +
  ggtitle("Contribution Count, by Candidate (2008)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

grid.arrange(pBar_1_2008)


pLine_1_2008 <- ggplot(data = subset(
                                subset(prez2008.normDate, contb_receipt_amt>0),
                                !is.na(contb_receipt_amt)
                                    ),
                       aes(x=contb_receipt_amt, 
                           y=..count..)
                       ) +
  geom_freqpoly(binwidth = 0.1) +
  ggtitle("Contribution Count, by Amount (2008)") +
  scale_x_log10(limits = c(1,10000))

grid.arrange(pLine_1_2008)
```
The basic univariate plots show the count of contributions made to each 
candidate, and the count of all contribution amounts. From the bar graph, 
Obama and Clinton are the top two contribution count receivers, with upwards 
of 40,000 contributions, with various others above 20,000. Obama has the most 
counts by far, but not as many as during his incumbent year of 2012. From the 
line plot, it appears that the distribution of contribution amounts is 
approximately normal or bimodal, after transforming the x-axis to logarithmic. 
Also, the highest contribution counts are for approximately $100 in size, with
another large peak at approximately $1,000.

_**Part 3.5. dplyr ops: get total contb receipt amt**_
Create a new data frame to get total contb receipt amt for 2012 Presidental 
Candidates. The new data frame has the following structure:
```{r 3_5 }
prez2008.normDate.cont_by_name <- prez2008.normDate %>%
  group_by(cand_nm) %>%
  summarise(sum_amnts = sum(contb_receipt_amt),
            n = n()) %>%
  arrange(cand_nm)

str(prez2008.normDate.cont_by_name)

###manual check
#by(prez2008.normDate$contb_receipt_amt, prez2008.normDate$cand_nm, sum)
#sum(prez2008.normDate$contb_receipt_amt)

```

The total itemized contributions for each candidate are:
```{r  }
###manual check
by(prez2008.normDate$contb_receipt_amt, prez2008.normDate$cand_nm, sum)
```

With the sum for all candidates:
```{r  }
sum(prez2008.normDate$contb_receipt_amt)
```

_**Part 3.6. More plots: distributions and sums of It. contribution amounts**_
```{r 3_6, fig.height=14 }
pBox_1_2008 <- ggplot(data = prez2008.normDate, 
                 aes(x = cand_nm, 
                     y = contb_receipt_amt)
                 ) +
  geom_boxplot() + 
  ylim(0, quantile(prez2008.normDate$contb_receipt_amt,0.9999)) + 
  ggtitle("Itemized Contribution Amount, by Candidate (2008)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

pBar_2_2008 <- ggplot(data = prez2008.normDate.cont_by_name, 
                 aes(x = cand_nm, 
                     y = sum_amnts)
                 ) +
  geom_bar(stat="identity") + 
  ggtitle("Sum of Itemized Contributions, by Candidate (2008)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))


#plot both 
grid.arrange(pBar_1_2008, pBox_1_2008, pBar_2_2008)
#
```
The first bar graph of each candidates total number of contributions is 
presented along with the box plot of the same candidate and the distribution 
of the contribution amounts. Additionally, from the data in the recently 
created data frame shows a bar graph of the total contribution amount each 
candidate has received.

Focusing on the larger amounts and larger contribution totals, Obama and 
Clinton all have high contribution counts. However, Obama and Clinton have a 
very different box plot distribution - Obama's contributions were all near 
$0 contribution amount (probably more like $100, from the previous line plot), 
but Clinton differs by having higher contribution amounts and more spread out 
distribution. Furthermore, Clinton has the highest total contribution amount. 
The discrepancy is again probably due to itemized vs. unitemized contributions,
but cannot add non-itemized contributions as bulk amount because unitemized is 
lumped for the whole 2 year period. Addition of the full 2 year amount could 
potentially skew results. Therefore, Steps 7 and 8 are skipped.

_**Part 3.9. Heat map of average itemized contribution amount **_  

A new data frame is made in order to process data in creating a heat map.   
```{r 3_9, fig.height=10}
### create a hap map: State vs. Candidate.
# organize data to long format: candidate, state, average amount, count
prez2008.normDate.cont_by_st <- prez2008.normDate %>%
  group_by(cand_nm, contbr_st) %>%
  summarise(avg_amnts = mean(contb_receipt_amt),
            n = n()) %>%
  ungroup() %>%
  arrange(cand_nm)

#head(prez2008.normDate.cont_by_st)

pHeat_1_2008 <- ggplot(aes(y = prez2008.normDate.cont_by_st$contbr_st, 
           x = prez2008.normDate.cont_by_st$cand_nm, 
           fill = prez2008.normDate.cont_by_st$avg_amnts),
  data = prez2008.normDate.cont_by_st) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(c("green", "red"))(100)) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

#plot
grid.arrange(pHeat_1_2008)
```

Again, heat map shows the average contribution amounts by state for each 
candidate. This can be very useful if we wanted to compare a candidate's 
state-averaged contributions to another.

At this point, the exploratory data analysis is complete and a rough, 
high-level understanding of the data and its patterns obtained.  

***
***

# 4. Univariate Section
## 4.1 Univariate Plots Section
For convenience sake, univariate plots are repeated below from sections 1.4, 
2.4, and 3.4 above, and modified as needed to aid with readibility. For 
example, the line graphs of "contb\_receipt\_amt" count are combined on one 
plot to aid in comparison.
```{r Univariate_plots_1, fig.height=14}
grid.arrange(pBar_1_2008, pBar_1_2012, pBar_1_2016, ncol=1) 
```

```{r Univariate_plots_2, fig.height=8}
ggplot() +
  geom_freqpoly(data = subset(
                              subset(prez2016, contb_receipt_amt>0),
                              !is.na(contb_receipt_amt)
                              ),
                aes(x=contb_receipt_amt,
                    y=..count..,
                    color = 'blue'),
                binwidth = 0.1
                ) +
  geom_freqpoly(data = subset(
                              subset(prez2012.normDate, contb_receipt_amt>0),
                              !is.na(contb_receipt_amt)
                              ),
                aes(x=contb_receipt_amt, 
                    y=..count..,
                    color = 'green'),
                binwidth = 0.1
                ) +
  geom_freqpoly(data = subset(
                              subset(prez2008.normDate, contb_receipt_amt>0),
                              !is.na(contb_receipt_amt)
                              ),
                aes(x=contb_receipt_amt, 
                    y=..count..,
                    color = 'red'),
                binwidth = 0.1
                ) +
  scale_colour_manual(
    name="Contribution Count by Amount (Election Year 2008, 2012, 2016)", 
    values=c("blue", "green", "red"), 
    labels=c("EY2016", "EY2012", "EY2008")
                      ) +
  scale_x_log10(limits = c(1,10000)) 
```

## 4.2 Univariate Analysis
Some observations regarding the dataset are provided, given the questions from 
the project template.

- What is the structure of your dataset?

The structure of the dataset is provided in Parts 1.1, 2.1, and 3.1 for each 
2016, 2012, and 2008 presidential campaign data sets. Generally, the structure 
contains 18 columns for key fields, and hundreds of thousands of rows, for 
each contribution received.

- What is/are the main feature(s) of interest in your dataset?

The main features of the datasets are the fields: "cand\_nm", "contbr\_st",
and "contb\_receipt\_amt".

- What other features in the dataset do you think will help support your 
investigation into your feature(s) of interest?

Because of the independent nature of the other fields in the dataset, not 
many other features can be used in support of the features of interest. It 
is true that some fields like "contbr\_city" provide more information and 
could supplement "contbr\_st", but given the amount of values (11036 
different entries for city), the clean up would be great in order to be 
handled on a laptop's computing capacity.

- Did you create any new variables from existing variables in the dataset?

Yes, variables that summed and averaged "contb\_receipt\_amt", the only 
numeric field, by state or candidate name were created. 

- Of the features you investigated, were there any unusual distributions? 
Did you perform any operations on the data to tidy, adjust, or change the
form of the data? If so, why did you do this?

The main unusual distribution occurred after seeing results in Part 1.6. 
Clinton's total seemed much greater than any other candidate, though the 
contribution counts were not as great. The discussion in that section 
explains the vital need for inclusion of unitemized contributions to paint 
the entire picture - only itemized contributions were given in this dataset. 

# 5. Bivariate Section
## 5.1 Bivariate Plots Section
For convenience sake, bivariate plots are repeated below from Parts 1.6, 2.6, 
and 3.6 above. 
```{r Bivariate_plots_1, fig.height=16}
grid.arrange(pBox_1_2008, pBox_1_2012, pBox_1_2016, ncol=1)
```

## 5.2 Bivariate Analysis
Some observations regarding the dataset are provided, given the questions 
from the project template.

- Talk about some of the relationships you observed in this part of the 
investigation. How did the feature(s) of interest vary with other features in 
the dataset? 

The box plot was the main tool used in comparing two variables, given that 
only 1 variable in the entire data set was numeric. The relationships and 
other understanding gained from the dataset from the bivariate plots is 
discussed in Parts 1.6, 2.6, and 3.6 above. 

- Did you observe any interesting relationships between the other features 
(not the main feature(s) of interest)?

There does appear to be a relationship between the distributions of grass 
root campaigns (Sanders, Cruz, etc.) and more establishment-supported 
campaigns (Cristie, Bush, etc.), though not formally quanified. In the 
grass root campaigns the means, medians, and range of amounts for majority 
of the contributions are much lower than the establishment campaigns. 

- What was the strongest relationship you found?

Unfortunately, because only one numeric variable was provided, no quantitative 
relationship parameters like Pearson's r coefficient could be calculated.

# 6. Multivariate Section
## 6.1 Multivariate Plots Section
For convenience sake, bivariate plots are repeated below from Parts 1.6, 2.6, 
and 3.6 above. 
```{r Multivariate_plots_1, fig.height=30}
grid.arrange(pHeat_1_2008, pHeat_1_2012, pHeat_1_2016, ncol=1)
```

## 6.2 Multivariate Analysis
Some observations regarding the dataset are provided, given the questions 
from the project template.

- Talk about some of the relationships you observed in this part of the 
investigation. Were there features that strengthened each other in terms of 
looking at your feature(s) of interest?

The heat map and scatterplot matrix were the main tools used in comparing 3 
variables, given that only 1 variable in the entire data set was numeric 
(other 2 fields had to be non-numeric, qualitative values). The relationships 
are first introduced in Parts 1.9, 2.9, and 3.9 above. No further insight is 
gained because state-level contributions is at a level of granularity that is 
above the scope of this exploratory data analysis.

- Were there any interesting or surprising interactions between features?

Some interesting, qualitative observations can be obtained. First, there 
seem to be non-states in the data set, such as "contbr\_st". This could be 
either a bug, or some proxy for a state or region that needs to be looked into
further. The red elements are those states that have contributed to a 
candidate highly. Likewise, those candidates with non-existant support 
from majority of states did not get very far in the election process, e.g. 
John H Cox or Thaddeus McCotter. 

- OPTIONAL: Did you create any models with your dataset? Discuss the strengths 
and limitations of your model.

Unfortunately, because only one numeric variable was provided, it would have 
been difficult to build a model, at least within the scope of this 
exploratory data analysis assignment, the tools and concepts available. 
This is certainly a step in a follow-up analysis of increased rigor.

***  

# 7. Final Plots and Summary

### Plot One  

```{r Plot_One, fig.width=8, fig.height=6}
ggplot() +
  geom_freqpoly(data = subset(
                              subset(prez2016, contb_receipt_amt>0),
                              !is.na(contb_receipt_amt)
                              ),
                aes(x=contb_receipt_amt,
                    y=..count..,
                    color = 'blue'),
                binwidth = 0.1
                ) +
  geom_freqpoly(data = subset(
                              subset(prez2012.normDate, contb_receipt_amt>0),
                              !is.na(contb_receipt_amt)
                              ),
                aes(x=contb_receipt_amt, 
                    y=..count..,
                    color = 'green'),
                binwidth = 0.1
                ) +
  geom_freqpoly(data = subset(
                              subset(prez2008.normDate, contb_receipt_amt>0),
                              !is.na(contb_receipt_amt)
                              ),
                aes(x=contb_receipt_amt, 
                    y=..count..,
                    color = 'red'),
                binwidth = 0.1
                ) +
  scale_colour_manual(
    name="Election Year [2008, 2012, 2016]",
    values=c("blue", "green", "red"), 
    labels=c("EY2016", "EY2012", "EY2008")
                      ) +
  ggtitle("Contribution Count by Amount (Election Year 2008, 2012, 2016)") +
  scale_x_log10(limits = c(1,10000)) +
  theme(legend.justification=c(0,1), legend.position=c(0,1))

```

### Description One  

The first chosen plot is a univariate comparison of the count of 
"cont\_receipt\_amt" by 3 election years. This plot is a first attempt at 
answering the question "how have contribution amounts varied over the last 
few election year cycles?". The general shape is approximately the same 
approximately normal distribution, with a potential bimodal distribution for 
EY2008. The counts of large contributions (>$1,000) have decreased from EY2008 
to EY2012 and EY2016 from about 50,000 counts to 35,000. Furthermore, only 
looking at EY2008 and EY2016 - the 2 non-incumbent years - EY2016 spikes at 
~$100 contribution amount at 70,000 counts, while EY2008 has two spikes at 
$200-$500 (50,000 counts) and $2,000-$5,000 (50,000 counts). An interpretation 
of this decrease in donation amounts from EY2008 to EY2016 is that the 
average, middle class population is starting to contribute to their candidates 
of choice, thus voting with their dollars. This is an important note in the 
face-off in Clinton vs. Sanders for the Democratic ticket - the unitemized 
contribution funding finding from the section above, combined with the shift 
in itemized contribution from this plot, and the box plot of contribution 
amount from the section above - may indicate that the population, generally
middle class by people count, is leaning in this direction. 

### Plot Two

```{r Plot_Two, fig.height=14}
##### modify data frame to fill in plots

### 2012
# 2016 Democrats
D2016 <- c("Clinton, Hillary Rodham", "Sanders, Bernard", 
           "O'Malley, Martin Joseph", "Lessig, Lawrence", 
           "Webb, James Henry Jr.")
# 2016 Republican
R2016 <- c("Carson, Benjamin S.", "Cruz, Rafael Edward 'Ted'", "Bush, Jeb", 
           "Rubio, Marco", "Trump, Donald J.", "Paul, Rand", "Fiorina, Carly", 
           "Kasich, John R.", "Walker, Scott", "Christie, Christopher J.", 
           "Graham, Lindsey O.", "Huckabee, Mike", "Jindal, Bobby", 
           "Perry, James R. (Rick)", "Santorum, Richard J.", 
           "Pataki, George E.")
prez2016$party <- ifelse(prez2016$cand_nm %in% D2016, 
                         "Dem.", 
                         ifelse(
                           prez2016$cand_nm %in% R2016, 
                           "Rep.", 
                           "Other")
                         )

### 2012
# 2012 Democrats
D2012 <- c("Obama, Barack")
# 2012 Republican
R2012 <- c("Bachmann, Michele", "Cain, Herman", "Gingrich, Newt", 
           "Huntsman, Jon", "McCotter, Thaddeus G", "Paul, Ron", 
           "Pawlenty, Timothy", "Perry, Rick", "Romney, Mitt", 
           "Santorum, Rick")
#2012 Libertarian
L2012 <- c("Johnson, Gary Earl")
#2012 Other
O2012 <- c("Roemer, Charles E. 'Buddy' III")
#2012 Green
G2012 <- c("Stein, Jill")

prez2012.normDate$party <- ifelse(
                            prez2012.normDate$cand_nm %in% D2012, 
                            "Dem.", 
                            ifelse(
                              prez2012.normDate$cand_nm %in% R2012, 
                              "Rep.", 
                              "Other")
                         )

### 2008
# 2008 Democrats
D2008 <- c("Obama, Barack","Clinton, Hillary Rodham","Edwards, John",
           "Richardson, Bill","Dodd, Christopher J",
           "Biden, Joseph R Jr","Kucinich, Dennis J","Gravel, Mike")
# 2012 Republican
R2008 <- c("McCain, John S","Romney, Mitt","Giuliani, Rudolph W",
           "Paul, Ron","Thompson, Fred Dalton","Huckabee, Mike",
           "Tancredo, Thomas Gerald","Brownback, Samuel Dale",
           "Hunter, Duncan","Thompson, Tommy G","Cox, John H",
           "Gilmore, James S III")

prez2008.normDate$party <- ifelse(
                            prez2008.normDate$cand_nm %in% D2008, 
                            "Dem.", 
                            ifelse(
                              prez2008.normDate$cand_nm %in% R2008, 
                              "Rep.", 
                              "Other")
                         )


## plot
pBox_1_2008_color <- ggplot(data = prez2008.normDate, 
                 aes(x = cand_nm, 
                     y = contb_receipt_amt, 
                     color = party)
                 ) +
  geom_boxplot() + 
  ylim(0, quantile(prez2008.normDate$contb_receipt_amt,0.9999)) + 
  ggtitle("Itemized Contribution Amount, by Candidate (2008)") +
  scale_colour_manual(name="Party", 
                      values=c("blue", "red", "grey"), 
                      labels=c("Dem.", "Rep.", "Other")
                      ) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

pBox_1_2012_color <- ggplot(data = prez2012.normDate, 
                      aes(x = cand_nm,
                          y = contb_receipt_amt,
                          color = party)
                      ) +
  geom_boxplot() + 
  ylim(0, quantile(prez2012.normDate$contb_receipt_amt,0.9999)) + 
  ggtitle("Itemized Contribution Amount, by Candidate (2012)") +
  scale_colour_manual(name="Party", 
                      values=c("blue", "grey", "red"), 
                      labels=c("Dem.", "Other", "Rep.")
                      ) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

pBox_1_2016_color <- ggplot(data = prez2016, 
                      aes(x = cand_nm, 
                          y = contb_receipt_amt,
                          color = party)
                      ) +
  geom_boxplot() + 
  ylim(0, quantile(prez2016$contb_receipt_amt,0.9999)) + 
  ggtitle("Itemized Contribution Amount, by Candidate (2016)") +
  scale_colour_manual(name="Party", 
                      values=c("blue", "red", "grey"), 
                      labels=c("Dem.", "Rep.", "Other")
                      ) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

# show plots
grid.arrange(pBox_1_2008_color, pBox_1_2012_color, pBox_1_2016_color, ncol=1)

```

### Description Two

The second chosen plot is a bivariate comparison of the box plots of 
"cont\_receipt\_amt" of each candidate by 3 election years (technically 
this is multivariate with the creation of the party, but this coloring was 
only meant to aid in elucidating the candidates). This plot is an attempt to 
show the evolution in time of the winning candidates along with their funding 
distribution. Tracking Obama starting in 2008 when he first won the election, 
the distribution is highly skewed towards the smaller values for the median. 
Clinton, however, is higher in median and significantly higher in the range of 
the IQR. This behavior is repeated in 2016, with Clinton being the constant. 
Although her median has dropped to that of Sanders, the quartile range between 
75th and 50th remains significantly higher than Sanders, which indicates a 
presence of wealthy donors. In my opinion, although the absolute amount 
donated remains important. Even if distribution favors the populist with 
skew towards the lower contribution amounts is very low total amount, the 
expectation is that this will not bring the name out and election will 
probably not happen - a critical mass is needed, with enough money to get the 
information out to voters across the country. Conversely, if the big money 
receipient receives massive amounts of money, this lack of concern for 
resources increases the ability to overwhelm the competition in terms of TV 
commercials, campaign events, data analysis teams and hardware, logistical 
support, etc. The takeaway from this plot suggests that Clinton-Sanders 
resembles a similar pattern to Clinton-Obama, with Sanders appearing on the 
winning side as Obama did.

### Plot Three

```{r Plot_Three, fig.width=6, fig.height=18}
##### obtain list of state names in abbr form
states <- c("AL", "MT", "AK", "NE", "AZ", "NV", "AR", "NH", "CA", "NJ", "CO", 
            "NM", "CT", "NY", "DE", "NC", "FL", "ND", "GA", "OH", "HI", "OK", 
            "ID", "OR", "IL", "PA", "IN", "RI", "IA", "SC", "KS", "SD", "KY", 
            "TN", "LA", "TX", "ME", "UT", "MD", "VT", "MA", "VA", "MI", "WA", 
            "MN", "WV", "MS", "WI", "MO", "WY")

prez2016.cont_by_st.clean <- subset(prez2016.cont_by_st, 
                                    prez2016.cont_by_st$contbr_st %in% states)
prez2012.normDate.cont_by_st.clean <- subset(
  prez2012.normDate.cont_by_st, 
  prez2012.normDate.cont_by_st$contbr_st %in% states)
prez2008.normDate.cont_by_st.clean <- subset(
  prez2008.normDate.cont_by_st, 
  prez2008.normDate.cont_by_st$contbr_st %in% states)

### plot
pHeat_1_2016_clean <- ggplot(aes(y = contbr_st,
                                 x = cand_nm, 
                                 fill = avg_amnts
                                 ),
                             data = prez2016.cont_by_st.clean
                             ) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(c("green", "red"))(100)) +
  ggtitle("Heat map of State Average Campaign Contribution (EY2016)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

pHeat_1_2012_clean <- ggplot(aes(y = contbr_st,
                                 x = cand_nm,
                                 fill = avg_amnts
                                 ),
                             data = prez2012.normDate.cont_by_st.clean
                             ) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(c("green", "red"))(100)) +
  ggtitle("Heat map of State Average Campaign Contribution (EY2012)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

pHeat_1_2008_clean <- ggplot(aes(y = contbr_st,
                                 x = cand_nm,
                                 fill = avg_amnts
                                 ),
                             data = prez2008.normDate.cont_by_st.clean
                             ) +
  geom_tile() +
  scale_fill_gradientn(colors = colorRampPalette(c("green", "red"))(100)) +
  ggtitle("Heat map of State Average Campaign Contribution (EY2008)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))

#plot all 
grid.arrange(pHeat_1_2008_clean, pHeat_1_2012_clean, pHeat_1_2016_clean, 
             ncol=1)
```

### Description Three

The third chosen plot is a multivariate comparison of the heat maps of 
"cont\_receipt\_amt" of each candidate and state by 3 election years. 
This plot is an attempt to show the evolution in time of the funding 
distribution mapped amongst the states. Besides a qualitative understanding of 
each candidate's state average funding performance and a comparison to their 
peers, there is not much information to pull from this. The one observation to 
note is the importance for a candidate to have support amongst all states - low 
country-wide support probably means no election win.

***  

# 8. Reflection

The successes of this investigation are contained in the exploratory data 
analysis step that led to many insights. Following the approach of the 
instructor in the example walk through of exploratory data analysis in Lesson 
6 of the course, a similar path was taken here: after obtaining and loading 
the data, steps were taken to get structure and summary information, 
scatterplot matrices to see how variables interacted with one another, and 
then more and more refined univariate, bivariate, and multivariate plots were 
created to get an understanding of the data. 

The difficulties were mostly not so much difficulties, but stepping stones and 
opportunities to understand R and the exploratory data analysis process better.
Some challenges were presented when a certain way the appearance of a plot was 
desired, but personal attempts were not fruitful; online investigation and 
consideration of other approaches allowed the solution to be obtain, and 
sometimes, even other insights were gained about R or the EDA process that 
were not anticipated or sought after.

The data has potentially revealed that count is an important indicator of 
potential presidency. For example, in Part 2.6., though Obama has lower median 
and lower total amount of funding, but he has many more counts than the field
and could potentially be used in prediction of the presidency. Furthermore, it
appears that the itemized count is an indicator, and unitemized amounts may not 
be necessary after all. Future work is to build a functional predictive model 
and see if these insights are correct. Further, it would be ideal if all 
contribution data, including the unitemized contributions that hindered 
progress on this investigation, could be included in one set for full analysis. 

***  


***
***


# Appendix 1. References.

Note: Rmd-formatted links removed and replaced with website URLs in order 
to fit 80 space margin. 

1. http://fec.gov/disclosurep/pnational.do

2. http://fec.gov/fecviewer/CandidateCommitteeDetail.do

3. http://fec.gov/fecviewer/CandCmteTransaction.do

4. Stackoverflow 
{e.g., https://stackoverflow.com/questions/1330989/rotating-
and-spacing-axis-labels-in-ggplot2; 
https://stackoverflow.com/questions/24932940/ggplot-does-not-show-legend-in-
geom-histogram; 
https://stackoverflow.com/questions/8005154/conditionally-remove-
dataframe-rows-with-r;  
https://stackoverflow.com/questions/15016723/how-to-add-column-into-a-
dataframe-based-on-condition-in-r-programming}

5. R documentation {via internal '?' and '??' commands}

6. https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

7. Miscellaneous 
{e.g., https://www.question-defense.com/2010/02/12/include-underscores-
while-using-markdown-without-them-being-bold-or-italic-indicators-for-text }

***

# Appendix 2. Comment Resolution from Reviews
Comments from review process included below and the way it was addressed is 
described below.

## Review 1 02/04/2016

https://review.udacity.com/#!/reviews/101761

**DOES NOT MEET SPECIFICATIONS**

### Code Functionality

SPECIFICATION
All code is functional (e.g. No Error is produced and RMD document is not 
prevented from bring knit.) 

MEETS SPECIFICATION  

_Reviewer Comments_
Code is functional, great job.

SPECIFICATION
The project almost never uses repetitive code where a function would be more 
ppropriate. The code references variables by name instead of using constants 
or column numbers.

MEETS SPECIFICATION

_Reviewer Comments_
Great job with the code!
To Exceed
Consider using custom functions to reduce repetitive code.
If interested check out this link: http://www.statmethods.net/management/
userfunctions.html
Requirements to exceed the specification 

_**Response:**_
Please see my note in the body about the repetitive code issue. I am aware of 
this, but due to the nature of this project (exploratory data analysis), 
additional work to code functions to reduce repetition would be unnecessary 
work - there are only 2 repeats of some portions of code: Repeating the code 
twice takes much less time than function generation and debugging that may 
need to happen.

### Project Readability

SPECIFICATION
All complex code is adequately explained with comments. It is always clear what 
the code is doing.

MEETS SPECIFICATION

_Reviewer Comments_
Good job commenting where necessary.
To Exceed
Consider removing commented blocks of code.
Requirements to exceed the specification 

_**Response:**_
Unnecessary comments removed to improve readibility.

SPECIFICATION
The code uses formatting techniques to improve code readability. All lines are 
shorter than 80 characters.

**DOES NOT MEET SPECIFICATION**

_Reviewer Comments_
Please refactor the code so that all lines of code are 80 characters or shorter.
As a suggestion, RStudio can draw a vertical line at 80 characters for you so 
that it's easy to check. This webpage shows how to do that with 'show margin'.
This is great coding, know that this habit is a standard for Google engineers. 
Imagine that! You are practising the same habit with some of the best engineers 
on Earth!
Requirements to exceed the specification 

_**Response:**_
Thank you for this comment, this really helped with the readibility of my code 
and write up. I will also keep this in mind for future projects, in both R, as 
well as any other programming languages that I use. 

SPECIFICATION
Markdown syntax is used in the code to improve readability of the knitted file.
There are no large sections of the knitted HTML file with bad readability.

MEETS SPECIFICATION

_Reviewer Comments_
Great job suppressing warning and code from the final report.
To Exceed
Consider using a global knitr chunk, like the following. This will allow plots 
to have constant dimensions and remove repetitive code from the script. Using 
Knitr global chunks is strongly encouraged, this demonstrates a deeper 
understanding of the R language and help increase readability or the final 
Knitted HTML document.
{r global_options, include=FALSE} knitr::opts_chunk$set(fig.width=12, 
fig.height=8, fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)
Requirements to exceed the specification 

_**Response:**_
Global knitr chunk enabled. I assume when I need to change the size of figures,
or some other change, I can overwrite the global settings with local ones. Will
implement and see how it goes!

### Quality of Analysis

SPECIFICATION
The project appropriately uses univariate, bivariate, and multivariate plots 
to explore most of the expected relationships in the data set.

EXCEEDS SPECIFICATION

_Reviewer Comments_
Excellent job with the analysis, the report includes a very unique structure 
were campaigns are explored per year. Then Univariate, Bivariate and 
Multivariate plots are presented.
Requirements to exceed the specification 

SPECIFICATION
Questions and findings are placed between blocks of R code regularly so it is 
clear what the student was thinking throughout the analysis.

MEETS SPECIFICATION

_Reviewer Comments_
Amazing job adding an introduction before any exploration begins.

SPECIFICATION
Reasoning is provided for the plots made throughout the analysis. Plots made 
follow a logical flow. Comments following plots accurately reflect the plots' 
contents. 

MEETS SPECIFICATION
 
SPECIFICATION
The project contains at least 20 visualizations. The visualizations are varied 
and show multiple comparisons and trends. Relevant statistics such as means, 
medians, quartiles, or confidence intervals are computed throughout the 
analysis when an inference is made about the data.

MEETS SPECIFICATION

_Reviewer Comments_
Great job with you visualization in general, good relationships were analice.
To Exceed
Consider adding more statistics that contribute to a deeper understanding of 
the data. It is recommended to have statistics like means and medians plotted 
in your graphs to improve visualization and comparison for readers. Please 
consider the following plot as an useful example:. Requirements to exceed
the specification 

SPECIFICATION
Visualizations made in the project depict the data in an appropriate manner 
that allows plots to be readily interpreted. Choices of plot type and 
parameters aid interpretability.

**DOES NOT MEET SPECIFICATION**

_Reviewer Comments_
I really enjoyed the visualizations, very professional and insightful. Please 
consider attending the following suggestions.
The following matrix is very hard to understand, since the axis tick labels 
are very hard to tell apart from each other. Please consider refactoring the 
plot so that it is more readable. I would suggest to plot less graphs in the 
matrix, pick out the most interesting relationships or variables.
For all plots like the following, consider that If the variable is ordinal, 
then an ordered color scheme should be chosen.
The color scheme should closely resemble the following:
Requirements to exceed the specification 

_**Response:**_
For bullet 1 (scatterplot matrix), in my opinion, this scatterplot matrix is 
sufficient because:
1) it is early on in the data set exploration, and not used in the Final Plots 
where refinement is required,
2) the variables of the data set are reduced to 5 key field already, from the 
original 18, 
3) because preliminary understanding about the data set is arbitrary and pretty 
subjective, personally this scatterplot matrix was useful for me and allowed me
to refine the relationship exploration via other univariate, bivariate, and 
multivariate plots. 

For bullet 2 (heat map), in my opinion, the variable of concern is not ordinal. 
The variable that is coloring the heat map is "avg_amnts", which is based off 
of a quantitative discrete/continuous variable of dollar amount, not a 
qualitative ordered species. Therefore, I believe the plotting set up is 
sufficient and correct.

Please advise on the two bullets above, and if perhaps the 
"Does Not Meet Specifications" for these two bullets can be reconsidered 
based on the explanation provided in this response.

### Final Plots and Summary

SPECIFICATION
The project includes a Final Plots and Summary section containing three plots 
and commentary. All plots in this section reflect what has been explored in 
the main body of the analysis.

MEETS SPECIFICATION
 
SPECIFICATION
The plots are well chosen and the plots fulfill at least 2 of the criteria. 
The plots are varied and reveal interesting trends and relationships.
Given criteria:
Draw comparisons.
Identify trends.
Engage a wide audience.
Explain a complicated finding.
Clarify a gap between perception and reality.
Enable the reader to digest large amounts of information.

EXCEEDS SPECIFICATION

_Reviewer Comments_
Excellent choices for the final plots, very insightful.
Requirements to exceed the specification 

SPECIFICATION
All plots have appropriately selected variables and are plotted in a way that 
accurately conveys the data/information (i.e findings in Final Plot 1 do not 
depend on the findings of Final Plot 2).

**DOES NOT MEET SPECIFICATION**

_Reviewer Comments_
Consider the advice suggested above in the "Quality of analysis" section for 
final plot three.

_**Response:**_
Please see my response to the provided suggestions in "Quality of analysis" 
section.

SPECIFICATION
All plots are labeled appropriately (axis labels, plot titles, axis units) 
and can be read and interpreted easily. Plots are scaled appropriately.

**DOES NOT MEET SPECIFICATION**

_Reviewer Comments_
Please consider adding tittles to all of the final plots.

_**Response:**_
Thanks for catching this oversight on my part. Titles added to plots, as needed.

SPECIFICATION
The reasoning and findings from each plot are explained and the text about 
each plot is descriptive enough to stand alone. Comments reflect the contents 
of the plots that they are associated with.

EXCEEDS SPECIFICATION

_Reviewer Comments_
Well done, rich plot descriptions were provided. Interesting findings and 
supporting analysis was drafted.
Requirements to exceed the specification 

### Reflection

SPECIFICATION
The project includes a Reflection section discussing the analysis performed.
 
MEETS SPECIFICATION
 
SPECIFICATION
The section reflects on how the analysis was conducted and reports on the 
struggles and successes throughout the analysis. The section provides at 
least one idea or question for future work.

EXCEEDS SPECIFICATION

_Reviewer Comments_
A great reflection that concludes an amazing analysis.
Requirements to exceed the specification 

_Additional Reviewer Comments_
Good start
A good first submission, a very high quality analysis has been drafted, there 
are some steps that must be taken in order to produce an even better final 
report.
Be certain that after the suggestions are addressed you will have an amazing 
final project and will be well on your way to becoming an excellent data 
analyst. Hope to see your submission soon, keep up the good work.

***
